---
title: "rbmi: Inference with Conditional Mean Imputation"
author: Alessandro Noci, Marcel Wolbers, Craig Gower-Page
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    number_sections: true
    citation_package: natbib
    base_format: rmarkdown::html_vignette
bibliography: "references.bib"
link-citations: true   
linkcolor: blue
pkgdown:
  as_is: true
vignette: >
  %\VignetteIndexEntry{rbmi: Inference with Conditional Mean Imputation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>"
)
```

# Introduction

As described in section 3.10.2 of the statistical specifications of the package (`vignette(topic = "stat_specs", package = "rbmi")`), two different types of variance estimators have been proposed for reference-based imputation methods in the statistical literature (@Bartlett2021). The first is the frequentist variance which describes the actual repeated sampling variability of the estimator and results in inference which is correct in the frequentist sense, i.e. hypothesis tests have accurate type I error control and confidence intervals have correct coverage probabilities under repeated sampling if the reference-based assumption is correctly specified (@Bartlett2021, @Wolbers2021). Reference-based missing data assumption are strong and borrow information from the control arm for imputation in the active arm. As a consequence, the size of frequentist standard errors for treatment effects may decrease with increasing amounts of missing data. The second is the so-called "information-anchored" variance which was originally proposed in the context of sensitivity analyses (@CroEtAl2019). This variance estimator is based on disentangling point estimation and variance estimation altogether. The resulting information-anchored variance is typically very similar to the variance under missing-at-random (MAR) imputation and increases with increasing amounts of missing data at approximately the same rate as MAR imputation. However, the information-anchored variance does not reflect the actual variability of the reference-based estimator and the resulting frequentist inference is highly conservative resulting in a substantial power loss. 

Reference-based conditional mean imputation combined with a resampling method such as the jackknife or the bootstrap was first introduced in @Wolbers2021. This approach naturally targets the frequentist variance. The information-anchored variance is typically estimated using Rubin's rules for Bayesian multiple imputation which are not applicable within the conditional mean imputation framework. However, an alternative information-anchored variance proposed by @Lu2021 can easily be obtained as we show below. The basic idea of @Lu2021 is to obtain the information-anchored variance via a MAR imputation combined with a delta-adjustment where delta is selected in a data-driven way to match the reference-based estimator. For conditional mean imputation, the proposal by @Lu2021 can be implemented by choosing the delta-adjustment as the difference between the conditional mean imputation under the chosen reference-based assumption and MAR on the original dataset. The variance can then be obtained via the jackknife or the bootstrap while keeping the delta-adjustment fixed. The resulting variance estimate is very similar to Rubin's variance. Moreover as shown in @CroEtAl2019, the variance of MAR-imputation combined with a delta-adjustment achieves even better information-anchoring properties than Rubin's variance for reference-based imputation.
Reference-based missing data assumptions are strong and borrow information from the control arm for imputation in the active arm.
This vignette demonstrates first how to obtain frequentist inference using reference-based conditional mean imputation using `rbmi`, and then shows that an information-anchored inference can also be easily implemented using the package. 

# Data and model specification

We use a publicly available example dataset from an antidepressant clinical trial of an active drug versus placebo. The relevant endpoint is the Hamilton 17-item depression rating scale (HAMD17) which was assessed at baseline and at weeks 1, 2, 4, and 6. Study drug discontinuation occurred in 24% of subjects from the active drug and 26% of subjects from placebo. All data after study drug discontinuation are missing and there is a single additional intermittent missing observation.

We consider an imputation model with the mean change from baseline in the HAMD17 score as the outcome (variable CHANGE in the dataset). The following covariates are included in the imputation model: the treatment group (THERAPY), the (categorical) visit (VISIT), treatment-by-visit interactions, the baseline HAMD17 score (BASVAL), and baseline HAMD17 score-by-visit interactions. A common unstructured covariance matrix structure is assumed for both groups. The analysis model is an ANCOVA model with the treatment group as the primary factor and adjustment for the baseline HAMD17 score. For this example, we assume that the imputation strategy after the ICE "study-drug discontinuation" is Jump To Reference (JR) for all subjects and the imputation is based on conditional mean imputation combined with jackknife resampling (but the bootstrap could also have been selected).

# Reference-based conditional mean imputation - frequentist inference

Conditional mean imputation combined with a resampling method such as jackknife or bootstrap naturally targets a frequentist estimation of the standard error of the treatment effect, thus providing a valid frequentist inference. Here we provide the code to obtain frequentist inference for reference-based conditional mean imputation using `rbmi`. 

The code used in this section is almost identical to the code in the quickstart vignette (`vignette(topic = "quickstart", package = "rbmi")`) except that we use conditional mean imputation combined with the jackknife (`method_condmean(type = "jackknife")`) here rather than Bayesian multiple imputation (`method_bayes()`). We therefore refer to that vignette and the help files for the individual functions for further explanations and details.  

## Draws {#draws}

We will make use of `rbmi::expand_locf()` to expand the dataset in order to have one row per subject per visit with missing outcomes denoted as `NA`. We will then construct the `data_ice`, `vars` and `method` input arguments to the first core `rbmi` function, `draws()`. Finally, we call the function `draws()` to derive the parameter estimates of the base imputation model for the full dataset and all leave-one-subject-out samples.

```{r draws_condmean}
library(rbmi)
library(dplyr)

dat <- antidepressant_data

# Use expand_locf to add rows corresponding to visits with missing outcomes to the dataset
dat <- expand_locf(
  dat,
  PATIENT = levels(dat$PATIENT), # expand by PATIENT and VISIT 
  VISIT = levels(dat$VISIT),
  vars = c("BASVAL", "THERAPY"), # fill with LOCF BASVAL and THERAPY
  group = c("PATIENT"),
  order = c("PATIENT", "VISIT")
)

# create data_ice and set the imputation strategy to JR for
# each patient with at least one missing observation
dat_ice <- dat %>% 
  arrange(PATIENT, VISIT) %>% 
  filter(is.na(CHANGE)) %>% 
  group_by(PATIENT) %>% 
  slice(1) %>%
  ungroup() %>% 
  select(PATIENT, VISIT) %>% 
  mutate(strategy = "JR")

# In this dataset, subject 3618 has an intermittent missing values which does not correspond
# to a study drug discontinuation. We therefore remove this subject from `dat_ice`. 
# (In the later imputation step, it will automatically be imputed under the default MAR assumption.)
dat_ice <- dat_ice[-which(dat_ice$PATIENT == 3618),]

# Define the names of key variables in our dataset and
# the covariates included in the imputation model using `set_vars()`
vars <- set_vars(
  outcome = "CHANGE",
  visit = "VISIT",
  subjid = "PATIENT",
  group = "THERAPY",
  covariates = c("BASVAL*VISIT", "THERAPY*VISIT")
)

# Define which imputation method to use (here: conditional mean imputation with jackknife as resampling) 
method <- method_condmean(type = "jackknife")

# Create samples for the imputation parameters by running the draws() function
drawObj <- draws(
  data = dat,
  data_ice = dat_ice,
  vars = vars,
  method = method,
  quiet = TRUE
)
drawObj
```

## Impute

We can use now the function `impute()` to perform the imputation of the original dataset and of each leave-one-out samples using the results obtained at the previous step. 

```{r}
references <- c("DRUG" = "PLACEBO", "PLACEBO" = "PLACEBO")
imputeObj <- impute(drawObj, references)
imputeObj
```

## Analyse

Once the datasets have been imputed, we can call the `analyse()` function to apply the complete-data analysis model (here ANCOVA) to each imputed dataset.

```{r}

# Set analysis variables using rbmi function "set_vars"
vars_an <- set_vars(
  group = vars$group,
  visit = vars$visit,
  outcome = vars$outcome,
  covariates = "BASVAL"
)

# Analyse MAR imputation with derived delta adjustment
anaObj <- analyse(
  imputeObj,
  rbmi::ancova,
  vars = vars_an
)
anaObj
```

## Pool

Finally, we can extract the treatment effect estimates and perform inference using the jackknife variance estimator. This is done by calling the `pool()` function.

```{r}
poolObj <- pool(anaObj)
poolObj
```

This gives an estimated treatment effect of 
`r paste(formatC(poolObj$pars$trt_7$est,format="f",digits=2)," (95% CI ", 
         formatC(poolObj$pars$trt_7$ci[1],format="f",digits=2), " to ",
         formatC(poolObj$pars$trt_7$ci[2],format="f",digits=2),")",sep="")`
at the last visit with an associated p-value of `r formatC(poolObj$pars$trt_7$pval,format="f",digits=3)`.

# Reference-based conditional mean imputation - information-anchored inference

In this section, we present how the estimation process based on conditional mean imputation combined with the jackknife can be adapted to obtain an information-anchored variance following the proposal by @Lu2021. 
## Draws

The code for the pre-processing of the dataset and for the "draws" step is equivalent to the code provided for the frequentist inference. Please refer to [that section](#draws) for details about this step. 

```{r, eval=FALSE}

<<draws_condmean>>

```

## Imputation step including calculation of delta-adjustment

The proposal by @Lu2021 is to replace the reference-based imputation by a MAR imputation combined with a delta-adjustment where delta is selected in a data-driven way to match the reference-based estimator. In `rbmi`, this is implemented by first performing the imputation under the defined reference-based imputation strategy (here JR) as well as under MAR separately. Second, the delta-adjustment is defined as the difference between the conditional mean imputation under reference-based and MAR imputation, respectively, on the original dataset.  

To simplify the implementation, we have written a function `get_delta_match_refBased` that performs this step. The function takes as input arguments the `draws` object, `data_ice` (i.e. the `data.frame` containing the information about the intercurrent events and the imputation strategies), and `references`, a named vector that identifies the references to be used for reference-based imputation methods. The function returns a list containing the imputation objects under both reference-based and MAR imputation, plus a `data.frame` which contains the delta-adjustment.

```{r}

#' Get delta adjustment that matches reference-based imputation
#' 
#' @param draws: A `draws` object created by `draws()`.
#' @param data_ice: `data.frame` containing the information about the intercurrent events and the imputation strategies. Must represent the desired imputation strategy and not the MAR-variant.
#' @param references: A named vector. Identifies the references to be used for reference-based imputation methods.
#' 
#' @return 
#' The function returns a list containing the imputation objects under both reference-based and MAR imputation, plus a `data.frame` which contains the delta-adjustment.
#' 
#' @seealso `draws()`, `impute()`.
get_delta_match_refBased <- function(draws, data_ice, references) {
  
  # Impute according to `data_ice`
  imputeObj <- impute(
    draws = drawObj,
    update_strategy = data_ice,
    references = references
  )
  
  vars <- imputeObj$data$vars
  
  # Access imputed dataset (index=1 for method_condmean(type = "jackknife"))
  cmi <- extract_imputed_dfs(imputeObj, index = 1, idmap = TRUE)[[1]]
  idmap <- attributes(cmi)$idmap
  cmi <- cmi[, c(vars$subjid, vars$visit, vars$outcome)]
  colnames(cmi)[colnames(cmi) == vars$outcome] <- "y_imp"
  
  # Map back original patients id since rbmi re-code ids to ensure id uniqueness
  
  cmi[[vars$subjid]] <- idmap[match(cmi[[vars$subjid]], names(idmap))]
  
  # Derive conditional mean imputations under MAR
  dat_ice_MAR <- data_ice 
  dat_ice_MAR[[vars$strategy]] <- "MAR"
  
  # Impute under MAR 
  # Note that in this specific context, it is desirable that an update   
  # from a reference-based strategy to MAR uses the exact same data for     # fitting the imputation models, i.e. that available post-ICE data are 
  # omitted from the imputation model for both. This is the case when    
  # using argument update_strategy in function impute(). 
  # However, for other settings (i.e. if one is interested in switching to
  # a standard MAR imputation strategy altogether), this behavior is  
  # undesirable and, consequently, the function throws a warning which 
  # we suppress here. 
  suppressWarnings(
    imputeObj_MAR <- impute(
      draws,
      update_strategy = dat_ice_MAR
    )
  ) 
  
  # Access imputed dataset (index=1 for method_condmean(type = "jackknife"))
  cmi_MAR <- extract_imputed_dfs(imputeObj_MAR, index = 1, idmap = TRUE)[[1]]
  idmap <- attributes(cmi_MAR)$idmap
  cmi_MAR <- cmi_MAR[, c(vars$subjid, vars$visit, vars$outcome)]
  colnames(cmi_MAR)[colnames(cmi_MAR) == vars$outcome] <- "y_MAR"
  
  # Map back original patients id since rbmi re-code ids to ensure id uniqueness
  cmi_MAR[[vars$subjid]] <- idmap[match(cmi_MAR[[vars$subjid]], names(idmap))]
  
  # Derive delta adjustment "aligned with ref-based imputation",
  # i.e. difference between ref-based imputation and MAR imputation
  delta_adjust <- merge(cmi, cmi_MAR, by = c(vars$subjid, vars$visit), all = TRUE)
  delta_adjust$delta <- delta_adjust$y_imp - delta_adjust$y_MAR

  ret_obj <- list(
    imputeObj = imputeObj,
    imputeObj_MAR = imputeObj_MAR,
    delta_adjust = delta_adjust
  )
  
  return(ret_obj)
}

references <- c("DRUG" = "PLACEBO", "PLACEBO" = "PLACEBO")

res_delta_adjust <- get_delta_match_refBased(drawObj, dat_ice, references)

```

## Analyse

We use the function `analyse()` to add the delta-adjustment and perform the analysis of the imputed datasets under MAR. `analyse()` will take as the input argument `imputations = res_delta_adjust$imputeObj_MAR`, i.e. the imputation object corresponding to the MAR imputation (and not the JR imputation). The argument `delta` can be used to add a delta-adjustment prior to the analysis and we set this to the delta-adjustment obtained in the previous step: `delta = res_delta_adjust$delta_adjust`.

```{r}

# Set analysis variables using rbmi function "set_vars"
vars_an <- set_vars(
  group = vars$group,
  visit = vars$visit,
  outcome = vars$outcome,
  covariates = "BASVAL"
)

# Analyse MAR imputation with derived delta adjustment
anaObj_MAR_delta <- analyse(
  res_delta_adjust$imputeObj_MAR,
  rbmi::ancova,
  delta = res_delta_adjust$delta_adjust,
  vars = vars_an
)
```

## Pool

We can finally use the `pool()` function to extract the treatment effect estimate (as well as the estimated marginal means) at each visit and apply the jackknife variance estimator to the analysis estimates from all the imputed leave-one-out samples.

```{r}

poolObj_MAR_delta <- pool(anaObj_MAR_delta)
poolObj_MAR_delta
```

This gives an estimated treatment effect of 
`r paste(formatC(poolObj_MAR_delta$pars$trt_7$est,format="f",digits=2)," (95% CI ", 
         formatC(poolObj_MAR_delta$pars$trt_7$ci[1],format="f",digits=2), " to ",
         formatC(poolObj_MAR_delta$pars$trt_7$ci[2],format="f",digits=2),")",sep="")`
at the last visit with an associated p-value of `r formatC(poolObj_MAR_delta$pars$trt_7$pval,format="f",digits=3)`. 
Per construction of the delta-adjustment, the point estimate is identical to the frequentist analysis. However, its standard error is much larger (`r formatC(poolObj_MAR_delta$pars$trt_7$se,format="f",digits=2)` vs. `r formatC(poolObj$pars$trt_7$se,format="f",digits=2)`). Indeed, the information-anchored standard error (and the resulting inference) is very similar to the results for Baysesian multiple imputation using Rubin's rules for which a standard error of `r 1.13` was reported in the quickstart vignette (`vignette(topic = "quickstart", package = "rbmi"`). Of note, as shown e.g. in @Wolbers2021, hypothesis testing based on the information-anchored inference is very conservative, i.e. the actual type I error is much lower than the nominal value. Hence, confidence intervals and $p$-values based on information-anchored inference should be interpreted with caution.

# References {.unlisted .unnumbered}